Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-29 Manuscript/TEM_draft.tex
Removed comments

\begin{document}
%\modulolinenumbers[1]     % Line numbering on every line

@@

%Gavin's general comment:I have made some comments on Thomas’ paper. I think it generally looks good. The one thing that I think might be tricky to get past reviewers is the comparison between the ML and Bayesian trees. The problem is that you can’t tell if the poorer performance of ML is because of the inference method or the difference between the substitution models - the ML uses an overparameterised model (GTR) compared to the generating model, whereas the Bayesian analyses do not. It might be possible to cite your way out of this but I wouldn’t be surprised if a reviewer asks for new analyses by either using different ML software that can handle the HKY model (e.g. Garli) or running the Bayesian with the GTR model (although that might not be adequate because if I remember correctly RAXML uses an approximation for gamma that is not the same as in other software).

@@

We find that the number of living taxa with morphological characters and the overall number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology. 
% NC: I think you misunderstood Nick's comment here. He meant that living > fossil but that you can't put the number of characters in here too. I've reverted back to the original as I'm not sure his solution is correct
Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework. Additionally, we find that Bayesian methods consistently outperform other tree inference methods. We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses.

%Hummm Nick's comment again - That's a point to make in the discussion maybe. But let's wait for Gavin comments...
%it's possible that adding dating information etc. in tip-dating makes Bayesian dated total-evidence analyses (ie tip-dating) better than non-dated Bayesian total-evidence analyses (typical total-evidence approach), because more information is being used, and because dates give some information about phylogenetic position.  In any case, the study wasn't about tip-dating
% NC: Simplest thing here is to just leave out the mention of tip dating. Will need to fix in discussion
% NC: Also Nick makes another sensible comment about *why* are living species more important. Check this is covered in discussion 
%TG: I think that's covered in \subsection{Individual effects of missing data parameters} - \subsubsection{Missing data for living taxa ($M_{L}$)}. (If no living taxa, the fossils cannot branch.) 
% NC: OK good

@@

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
%G: Alan Lemmon has done a fair amount of work on missingness - worth checking out the work from his lab, e.g. http://www.evotutor.org/LemmonLab/Pdfs/Lemmon-etal2009.pdf TG: added reference (that was actually in the 2014 draft)
% NC: OK. But be careful here. When people give you comments you want to take them seriously and actually read the papers they suggest rather than just adding references. Are you sure Lemmon is just looking at molecular stuff? Think Gavin's point might be that he's done stuff on missingness in morphological data too which at present we don't really cite or cover at all. TG: I don’t like arguing on comments but this paper is about missing data in molecular matrices (still relevant though). The reason I removed it from the old version is that it states something kind of similar than all the others: overlap but it is specially important in bigger matrices where missing data can weaken the support in certain parts of the tree. The idea being probably close to the tree terrace paper.
% NC: Again let's chat tomorrow
Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods.
%TG: I swapped the results to be consistent, ML and MC first and then Bayesian % NC: Cool
We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the "best" tree topology more so than minimizing the amount of missing data in the fossil record. Additionally, we find that the ability of Total Evidence methods to recover the "best" tree topology is increased when using Bayesian methods.

@@

\subsection{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}.
%G:I don’t think there is a problem with this split but I expect reviewers might ask for justification. Is this typical of existing TE studies? TG: avoiding bias comment.
% NC: I think you misunderstood Gavin's point. He knows why you did it like this, but if usually TE studies are 95% living and 5% fossil (for example) then your results aren't really viable in real world. He ideally wants a statement like: this is a typical split between fossil and living tax across empirical TE studies (cite papers). However, I doubt this is the case. Actually aren't TE studeis usually dominated by fossil taxa?
% TG: Really depends. In the 5 I keep citing, Pyron is close to 50/50, Ronquist and Slater are more towards living and Beck is more towards fossil
% NC: Cool this is better and closer to what Gavin meant. I've condensed it a bit
Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \citealt{beckancient2014}, 58\% \citealt{schragocombining2013}, 48\% \citealt{pyrondivergence2011}, 31\% \citealt{ronquista2012} and 31\% \citealt{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa.
We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

To measure the support for each branch in our simulated phylogenies
%G: Do you mean you tested whether the simulated data generally had any phylogenetic signal with which to recover a tree? I was a bit unsure but mainly because I am hardwired to thinks of Pagel’s lambda when anyone mentions phylogenetic signal…
% NC: Haha that is exactly what I've said about a million times to you.
% TG: And after the literature review for the next paper (missing living in mammals) on community structure & Co. I start to get where you guys are comming for ;). % NC :P
we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

@@

\subsection{4. Comparing topologies}
We compared the topology of the "missing-data" trees to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology. We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa (i.e. taxa that frequently change position in different trees; \citealt{kearneyfragmentary2002}).
%G: Why these two metrics? Might need a few sentences to justify them in addition to the explanation of what they are that you already have.
%We used these metrics because they are widely used in literature \citep[][e.g.]{Bogdanowicz2012}, their relative performance is well described \citep{Kuhner04112014} % NC: Is that what you mean. "they relative performance" means nothing to me. 
%and because they illustrate two different aspects of tree topology (see Discussion). TG: ok I've fixed that towards: "as said above, they describe two different things but also they're cool because we know how they behave and the code is available to calculate them." - a bit more honest. % NC: This is better
We used these two metrics because they illustrate two different aspects of tree topology (see Discussion) but also because their performance in measuring differences in topology is well described \citep{Kuhner04112014} and well implemented \citep{Bogdanowicz2012}.
We normalised both metrics using methods described in \citet{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa. These metrics are described in detail below.

@@

\begin{landscape}
\begin{table}[ht]
\caption{
%G: I think this table needs to be decoded a bit. I wasn’t really sure what it was supposed to be getting across. Maybe it is because I am not familiar with the Bhatt… distance but I suspect that many readers will be similarly unfamiliar. TG: fixed clarity
Bhattacharyya Coefficients of the pairwise method comparisons, each of which corresponds to the normalised distance between the "best" tree and the "missing data" trees using either the Normalised Robinson-Foulds distance (RF) or the Normalised Triplets distance (Tr). Each line summarizes the distribution of the probability of overlap between pairs of tree inference methods. The values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the "correct" tree topology. % NC: check this is right, I think that's what Gavin wanted to be added - a way of interpreting what is going on
}
\centering
\begin{tabular}{lccccccc}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    Maximum Likelihood \textit{vs.} Bayesian consensus                 & $RF$ & 0.00 & 0.00 & 0.10 & 0.20 & 0.32 & 1.00 \\ 
                                                                       & $Tr$ & 0.34 & 0.49 & 0.61 & 0.62 & 0.75 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.03 & 0.54 & 0.69 & 0.64 & 0.77 & 0.98 \\ 
                                                                       & $Tr$ & 0.08 & 0.57 & 0.65 & 0.64 & 0.73 & 0.82 \\ 
    Maximum Likelihood \textit{vs.} Bayesian posterior trees           & $RF$ & 0.02 & 0.74 & 0.80 & 0.79 & 0.89 & 0.98 \\ 
                                                                       & $Tr$ & 0.21 & 0.67 & 0.73 & 0.72 & 0.77 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.00 & 0.00 & \textbf{0.00} & \textbf{0.01} & 0.01 & 0.04 \\ 
                                                                       & $Tr$ & 0.08 & 0.38 & 0.59 & 0.57 & 0.73 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Bayesian posterior trees           & $RF$ & 0.00 & 0.00 & \textbf{0.01} & \textbf{0.02} & 0.04 & 0.11 \\ 
                                                                       & $Tr$ & 0.21 & 0.36 & 0.56 & 0.55 & 0.74 & 0.87 \\ 
    Bayesian posterior tree \textit{vs.} Maximum Likelihood bootstraps & $RF$ & 0.50 & 0.77 & \textbf{0.85} & \textbf{0.85} & 0.96 & 1.00 \\ 
                                                                       & $Tr$ & 0.91 & 0.96 & \textbf{0.98} & \textbf{0.97} & 0.99 & 1.00 \\ 
   \hline
\end{tabular}
\label{Tab_Results-Difference_methods}
\end{table}
\end{landscape}

>>>>

\begin{document}

@@

@@

We find that the number of living taxa with morphological characters and the overall number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology. Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework. Additionally, we find that Bayesian methods consistently outperform other tree inference methods. We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses.

@@

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
%G: Alan Lemmon has done a fair amount of work on missingness - worth checking out the work from his lab, e.g. http://www.evotutor.org/LemmonLab/Pdfs/Lemmon-etal2009.pdf TG: added reference (that was actually in the 2014 draft)
% NC: OK. But be careful here. When people give you comments you want to take them seriously and actually read the papers they suggest rather than just adding references. Are you sure Lemmon is just looking at molecular stuff? Think Gavin's point might be that he's done stuff on missingness in morphological data too which at present we don't really cite or cover at all. TG: I don’t like arguing on comments but this paper is about missing data in molecular matrices (still relevant though). The reason I removed it from the old version is that it states something kind of similar than all the others: overlap but it is specially important in bigger matrices where missing data can weaken the support in certain parts of the tree. The idea being probably close to the tree terrace paper.
% NC: Again let's chat tomorrow
Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix). We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the "best" tree topology more so than minimizing the amount of missing data in the fossil record. Additionally, we find that the ability of Total Evidence methods to recover the "best" tree topology is increased when using Bayesian methods.

@@

\subsection{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \citealt{beckancient2014}, 58\% \citealt{schragocombining2013}, 48\% \citealt{pyrondivergence2011}, 31\% \citealt{ronquista2012} and 31\% \citealt{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

@@

\subsection{4. Comparing topologies}
We compared the topology of the "missing-data" trees to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology. We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa (i.e. taxa that frequently change position in different trees; \citealt{kearneyfragmentary2002}). We used these two metrics because they illustrate two different aspects of tree topology (see Discussion) but also because their performance in measuring differences in topology is well described \citep{Kuhner04112014} and well implemented \citep{Bogdanowicz2012}. We normalised both metrics using methods described in \citet{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa. These metrics are described in detail below.

@@

\begin{landscape}
\begin{table}[ht]
\caption{Bhattacharyya Coefficients of the pairwise method comparisons, each of which corresponds to the normalised distance between the "best" tree and the "missing data" trees using either the Normalised Robinson-Foulds distance (RF) or the Normalised Triplets distance (Tr). Each line summarizes the distribution of the probability of overlap between pairs of tree inference methods. The values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the "correct" tree topology.}
\centering
\begin{tabular}{lccccccc}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    Maximum Likelihood \textit{vs.} Bayesian consensus                 & $RF$ & 0.00 & 0.00 & 0.10 & 0.20 & 0.32 & 1.00 \\ 
                                                                       & $Tr$ & 0.34 & 0.49 & 0.61 & 0.62 & 0.75 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.03 & 0.54 & 0.69 & 0.64 & 0.77 & 0.98 \\ 
                                                                       & $Tr$ & 0.08 & 0.57 & 0.65 & 0.64 & 0.73 & 0.82 \\ 
    Maximum Likelihood \textit{vs.} Bayesian posterior trees           & $RF$ & 0.02 & 0.74 & 0.80 & 0.79 & 0.89 & 0.98 \\ 
                                                                       & $Tr$ & 0.21 & 0.67 & 0.73 & 0.72 & 0.77 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Maximum Likelihood bootstraps      & $RF$ & 0.00 & 0.00 & \textbf{0.00} & \textbf{0.01} & 0.01 & 0.04 \\ 
                                                                       & $Tr$ & 0.08 & 0.38 & 0.59 & 0.57 & 0.73 & 0.84 \\ 
    Bayesian consensus \textit{vs.} Bayesian posterior trees           & $RF$ & 0.00 & 0.00 & \textbf{0.01} & \textbf{0.02} & 0.04 & 0.11 \\ 
                                                                       & $Tr$ & 0.21 & 0.36 & 0.56 & 0.55 & 0.74 & 0.87 \\ 
    Bayesian posterior tree \textit{vs.} Maximum Likelihood bootstraps & $RF$ & 0.50 & 0.77 & \textbf{0.85} & \textbf{0.85} & 0.96 & 1.00 \\ 
                                                                       & $Tr$ & 0.91 & 0.96 & \textbf{0.98} & \textbf{0.97} & 0.99 & 1.00 \\ 
   \hline
\end{tabular}
\label{Tab_Results-Difference_methods}
\end{table}
\end{landscape}
